---
title: "Introduction to R"
output: 
  learnr::tutorial:
    css: css/style.css
    theme:
      version: 3
      bootswatch: sandstone
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(bslib)

if (Sys.getenv("CONTAINERIZED") == "TRUE") {
  source("/usr/bin/learnr/util.R")
} else {
  source(rprojroot::is_rstudio_project$find_file("interactive-tutorials/R/util.R"))
}

knitr::opts_chunk$set(echo = FALSE)
```

## Why Choose R?

Let me start with the most convincing argument I have ever seen for choosing
to use R (or, at least, some programming language) for data manipulation
and analysis:

```
marketValue=IF(AND(N212="BOH",N213="BOH")=TRUE,C213*P212,IF(AND(N212="Sell",N213="Sell")=TRUE,Q212,IF(AND(N212="BOH",N213="Sell")=TRUE,P212*C213,IF(AND(N212="Sell",N213="BOH")=TRUE,Q212))))
```
<small>Example from [this](https://stackoverflow.com/questions/11317001/converting-excel-functions-into-r) StackOverflow question.</small>

Let's be honest: If you've been using spreadsheets for any appreciable amount 
of time, this is probably not the *worst* formula you've ever seen. In fact, I 
find it pretty representative of the types of things you might find in a 
decently advanced spreadsheet. That's **before** you start trying to do 
calculations with dates...

Now, how would this look in R?

```{r eval=FALSE, include=TRUE, echo=TRUE}
#' If the `N` on the current row is "BOH" ("buy or hold"), assign 
#' the current market value (`Q`) to the `marketValue` column, otherwise, 
#' assign the current number of shares (`P`) times the next row's closing 
#' price (`C`).
table$marketValue <- with(table, ifelse(N == "BOH", Q, P * lead(C)))
```

<small><i>Those first few lines that start with `#` are **comments**. They're
only there to help explain what the script is doing. The computer ignores them.
Comments are great to (a) explain your reasoning to other people who look at 
your code and (b) help you remember what you were doing when you look back at 
your code in 6 months and need to change it slightly.</i></small>

So: Did you notice from the spreadsheet formula that the value for N213 doesn't
actually matter? If not, you're not alone. That formula is objectively hard to 
read.

So, what does this demonstrate? To be sure, spreadsheets (and other, more 
sophisticated **B**usiness **I**ntelligence tools) are extremely useful and a
complete data analysis ecosystem probably includes spreadsheets, BI tools, 
**and** a robust programming language. Each has their strengths and weaknesses. 
For R, the strengths include:

- **Scale**: You can work with much larger datasets much more quickly in R, 
since many of the operations are actually implemented in the very fastest 
languages available under the hood (like C and FORTRAN).
- **Access**: It's very straightforward to connect to and analyze data contained
in a database directly. This is also possible with spreadsheets, but once you
understand how to do it programmatically, it's considerably more user friendly.
- **Reproducibility**: Once you have written a script, you can use that script
100 or 100,000 times and expect to get the same result each time, so long as
your data doesn't change in strange ways. And, if your code is written clearly,
you can always look back and easily tell what you did every step of the way.
The availability of version control systems like Git ensure you can also inspect
how your code has changed over time and that you can produce a 
"similar-but-not-the-same" report as the one you did six months ago with
minimal fuss and confusion.
- **Complexity**: You can just do more with a script than you can in a 
spreadsheet. Simple linear regression, sure, use a spreadsheet. Combine multiple
tables before running a principle components analysis? Use a script. (Honestly,
combining multiple tables always feels easier to to in a script, regardless of
what you do after that.) Often, scripts can be a part of a workflow involving 
transforming data from its raw state to something that can be used/visualized 
in a spreadsheet or BI tool.
- **Interactivity**: Programming languages can be used to write programs, too.
Again, many BI tools offer interactivity in some form or another, but if you 
want to do more (or just different) than what the BI tool or spreadsheet
software supports, you'll need to write it yourself. There is a lot of support
in R for sharing interactive data analytics, charts, tools, etc.
- **Cost**: R is free. RStudio, the premier R development environment is also
free. All the R packages on the **C**omprehensive **R** **A**rchive **N**etwork 
(CRAN), they're free too. A huge library of training and learning materials, 
including some of the best books on learning R?[^1] You guessed it: freely 
available on the internet.

[^1]: Additional Resources: [R for Data Science](https://r4ds.had.co.nz/); 
[Advanced R](https://adv-r.hadley.nz/); [R Packages](https://r-pkgs.org/); 
[R Graphics Cookbook](http://www.cookbook-r.com/Graphs/); 
[Exploratory Data Analysis with R](https://leanpub.com/exdata); 
[Software Carpentry with R](http://swcarpentry.github.io/r-novice-inflammation/); 
[Hands on Programming with R](https://rstudio-education.github.io/hopr/); 
[Even More!](https://rstudio.com/resources/books/)


The best part is, if you're already familiar with using spreadsheets, looking
at tabular data (data in tables) or structured data (think, XML or JSON), and
thinking analytically, you are more than ready to quickly pick up R for 
basic data analysis and start growing your skills.

Have I convinced you yet? If so, go ahead and install R and RStudio[^2], 
and move on to the next section of this tutorial.

[^2]: [Install R/RStudio](https://rstudio-education.github.io/hopr/starting.html)

## Starting With Data

These tutorials are designed for folks who are already familiar (at least a 
little bit) with viewing, manipulating, and analyzing data. Whether your
experience is primarily with spreadsheets, SQL databases, or a business 
intelligence tool, you have all the foundation you need to 'level up` your
skills and pick up a real-live, full-fledged programming language whose primary
purpose is to speed up and streamline data analysis.

Since we're all data people here, let's start with a familiar topic: data! 
Specifically, how data is manipulated and described in R. As
you probably know (or are about to learn for the first time), most programming 
languages store data in special 'containers' called `r def_link("variables")`. 
Variables are typically given names that help you remember what type of 
information they contain. One of the most important (and frequent) things you 
will do as a programmer is perform `r def_link("operations")` on or with 
variables. There are lots of different operations (you already know a few), 
we'll examine a number of the operations available in R and in R packages later 
on. The images below demonstrate the use of variables and variable operations.

```{r fig1, results="asis"}
insert_figure(
  "01_store_variable.png", 
  "storing value in a variable", 
  caption = "Figure 1: Storing the value `15` in the variable `score`.", 
  max_width = "400px"
)
```

```{r fig2, results="asis"}
insert_figure(
  "02_variable_addition.png",
  "adding to a variable",
  caption = "Figure 2: Adding `10` to the value stored in `score` yields `25`."
)
```

The way you would express that in code is:

```{r echo=TRUE}
score <- 15
score + 10
```

Notice the block of text with the gray background above? That's a code block,
and it contains (you guessed it!) code. Below that, in the second box, is the
`r def_link("output")`, or the result, of the last `r def_link("expression")` 
in the code block, `score + 10`. Remember this format, you will see it 
throughout these tutorials.

Also, we've just used our first bit of R-specific `r def_link("syntax")`. If 
you've ever seen any computer code before, you've probably seen variable 
`r def_link("assignment")` written something like `name = value`, using 
the `=` sign as the assignment `r def_link("operator")`. Well, to be honest, 
that works in R, too, but R is special in that `x = 5`, `x <- 5`, and `5 -> x` 
all mean the same thing. Most R folks use that second form though, and that's 
the convention we'll adopt in these tutorials. (Hint: If you're using RStudio,
you can type `<-` with the `Alt+-` hotkey. It'll quickly become second nature.)

<small>You can get a listing of the basic arithmetic operators in R by typing 
'?Arithmetic' into the R console and the basic comparison operators by typing '?Comparison'.</small>

### Naming Variables

In R, like most programming languages, there are certain constraints around what
constitutes a 'valid' name for a variable. The rules are as follows:

1. Variable names can only include letters, digits, periods (.) and 
underscores (_).
2. A variable name must start with a letter or a period. If it starts with a 
period, the second character must be a letter.
3. Reserved words in R cannot be used as variable names. Reserved words include
'if', 'else', 'repeat', 'while', 'function', 'for', 'in', 'next', 'break', 
'TRUE', 'FALSE', 'NULL', 'Inf', 'NaN', 'NA', 'NA_integer_', 'NA_real_', 
'NA_complex_', 'NA_character_'. 

### Exercise 

*Try the following exercises in this interactive code block*

- Write the R code required to add two plus two
- Assign a value to a variable and do some math with it
- See if you can guess what symbols (operators) to use for arithmetic functions
- Bonus points if you can figure out addition, subtraction, multiplication,
division, and exponents
- Try to assign a value to a variable with an invalid name.

```{r exercise_one, exercise=TRUE, exercise.lines = 5, exercise.eval = T}
score <- 15
score + 10
```

```{r def_variables, context="server"}
observeEvent(input$`variables`, {
  showModal(modalDialog(
    title = "Variable",
    "In computer programming, a variable is a storage location paired with an associated symbolic name (an identifier) which contains a value. In other words, a variable is a storage location for data."
  ))
})
```

```{r def_operations, context="server"}
observeEvent(input$`operations`, {
  showModal(modalDialog(
    title = "Operation",
    "A process in which a number, quantity, expression, etc., is altered or manipulated according to formal rules, such as those of addition, multiplication, and differentiation. In programming, operations are not just limited to arithmetic, but can include examples such as rearrangement, adding or removing values to/from lists, writing data to a file, etc."
  ))
})
```

```{r def_output, context="server"}
observeEvent(input$`output`, {
  showModal(modalDialog(
    title = "Output",
    "Information that is processed by and sent 'out' from a computer or other electronic device. An example of output is 
anything viewed on your computer monitor screen, such as the words you type on your keyboard. Also, the result of an **operation**."
  ))
})
```

```{r def_expression, context="server"}
observeEvent(input$`expression`, {
  showModal(modalDialog(
    title = "Expression",
    "An expression is a syntactic entity in a programming language that may be evaluated to determine its value. It is a combination of one or more constants, variables, functions, and operators that the programming language interprets and processes."
  ))
})
```

```{r def_syntax, context="server"}
observeEvent(input$`syntax`, {
  showModal(modalDialog(
    title = "Syntax",
    "The set of rules that defines the combinations of symbols that are considered to be correctly structured statements or expressions in that language."
  ))
})
```

```{r def_assignment, context="server"}
observeEvent(input$`assignment`, {
  showModal(modalDialog(
    title = "Assignment",
    "An assignment statement sets and/or re-sets the value stored in the storage location(s) denoted by a variable name; in other words, it copies a value into the variable."
  ))
})
```

```{r def_operator, context="server"}
observeEvent(input$`operator`, {
  showModal(modalDialog(
    title = "Operator",
    "A character or characters that determine the action that is to be performed or considered."
  ))
})
```

## Data Types

When we talk about how data is stored in R, either in variables or in files, we
need to consider `r def_link("data types")`, or, what 'kind' of information
is in our variable. This is important because while the result of the operation
`1 + 1` should be pretty intuitive, the result of `'a' + 1` probably isn't. It
actually doesn't make sense to R, either, and if you try it you'll get the
following helpful message

```{r echo=FALSE, warning=TRUE}
tryCatch('a' + 1, error = function(e) cat(as.character(e)))
```

That is a genuine, grade-A error message, and you're going to see a lot of those
if you pursue the path of a programmer, especially at first. Error messages
don't really go away as you get more experience either, but your relationship
with them is likely to move from frustration to appreciation, since, with
practice, they can become extremely helpful in writing good, working code.

Ok, back to *data types*. In R, there are six basic data types:

- **logical**: `TRUE` or `FALSE`, also called a 'boolean' value
- **integer**: whole numbers, like `1`, `2`, `356`
- **double**: decimal numbers, like `1.5`, `2.0`, `3.9`
- **character**: letters and other alphanumeric characters, 
like `'a'`, `'@'`, `'1'`, `'cheese'`
- **complex**: complex numbers, like `1+2i`, `5+9i`
- **raw**: raw bytes represented in hexadecimal, like `c8`, `05`, `ff`

Some notes about data types:

- Of the six data types, **complex** and **raw** are likely to be less useful
day-to-day (excepting some more specialized use-cases), so we won't talk about
those much. In addition, **integer** is only there to allow communication 
between R and other languages that use *single-precision* data types, like C
or FORTRAN. There's a very high chance you won't need to try to use **integers**
directly.
- **double** data is the type for numbers you will work with most often. It's
what numbers that *look* like integers will be stored as, since `typeof(5)`
will return `"double"`. This means operations like `5 + 1.9` will just work,
since `5` and `1.9` are both of type `double`. 
- **character** data is denoted by quotes, either single `'` or double `"`. It 
doesn't matter much which one you use, so long as you're consistent. In this 
tutorial, you will see **character** data indicated by single quotes.

One last note, you can always check the data type of a value or variable using
the `typeof()` `r def_link("function")`. We'll talk (a lot) more about
functions, but for now, just know that running this R code, `typeof(x)`, would
provide you with the name of the data type stored in the variable `x`.

```{r def_data_types, context="server"}
observeEvent(input$`data-types`, {
  showModal(modalDialog(
    title = "Data Types",
    "An attribute of data which tells the computer how the programmer intends to use the data. A data type indicates which operations that can be done on the data, the meaning of the data, and the way values of that type can be stored."
  ))
})
```

```{r def_function, context="server"}
observeEvent(input$`function`, {
  showModal(modalDialog(
    title = "Function",
    "A block of organized, reusable code that is used to perform a single, related action. Functions typically take values to operate on, called 'arguments', and may or may not provide a value back to the code that invoked the function, called a 'return value'."
  ))
})
```

### Exercise 1

Satisfy yourself that you can store logical, double, and character
values in a variable and confirm the 'type of' the variable. Remember, you
can name a variable anything you want.

```{r exercise_two, exercise=TRUE, exercise.lines = 5, exercise.eval = T}
logical_value <- TRUE
typeof(logical_value)
```

### Data Type Conversion

In R, you will want to pay particular attention to the data types of your values
and variables, because R is a **dynamically-typed** language, meaning the types
of your values, variables, arguments, etc. will only be checked when your program
is run. This is in contrast to **statically-typed** languages that will check
types *before* running (or compiling) code and alert you to data type issues 
through error messages. Luckily, R provides you with the tools you need to 
ensure your variables are the type you are expecting and to `r def_link("cast")`
your variables to the appropriate type when you need to.

The `is.*()` family of functions are similar to the `typeof()` function, except
they allow you to confirm whether or not a given variable is a particular 
data type. There are `is.*()` functions for all the primitive data types and
for a lot of the more complex data types (or structures) we'll look at in future
tutorials. `is.double()`, for example, will return `TRUE` if the given argument
is a double type value, and `FALSE` if it is not. There are also `is.integer()`, 
`is.character()`, `is.complex()`, `is.logical()`, `is.raw()`, and `is.numeric()`.
`is.numeric()` is a bit of a special case, it will return `TRUE` for any value 
that can be interpreted as a number, meaning both integers and doubles.

The `as.*()` family of functions will attempt to convert (or cast) their 
argument to the given data type. There are also `as.*()` functions for all the
primitive data types, too. For example, `as.double("1")` will return the 
double value of `1.0`. There are also `as.integer()`, `as.character()`, 
`as.complex()`, `as.logical()`, `as.raw()`, and `as.numeric()`. `as.numeric()`
just converts arguments to doubles, so it is essentially the same as `as.double()`.

### Exercise 2

Satisfy yourself that you can check the types of variables and convert variables
from one type to another. Uncomment one or more lines below to see the result.

```{r exercise_three, exercise=TRUE, exercise.lines = 8, exercise.eval = T}
double_value <- 1.5
# is.character(double_value)
# is.double(double_value)

character_value <- as.character(double_value)
# is.character(character_value)
# is.double(character_value)
```

```{r def_cast, context="server"}
observeEvent(input$`cast`, {
  showModal(modalDialog(
    title = "Cast",
    span(
      span("Type casting is when you assign a value of one primitive data type to another type, such as converting the string value of "),
      tags$code("\"1\""),
      span(" to the numeric value "),
      tags$code("1"),
      span(".")
    )
  ))
})
```

## Control Flow

Since R is a real-live programming language, you'll want to think about **control flow**
when you're writing R code. Put simply, control flow is the order in which the
code you've written is evaluated. As the person writing that code, you can make
choices about that order. In general, there are two primary tools used to control
the flow of your program: choices and loops. 

### Choices - `If` Statements

In your code, you can make a *choice* about which code to run (or not run) based
on certain `r def_link("conditions")`. In R, choices take the form of `if` 
statements, calls to the `ifelse()` function, or the `switch()` function.

```{r def_condition, context="server"}
observeEvent(input$`conditions`, {
  showModal(modalDialog(
    title = "Condition",
    "In programming, a *condition* or *conditional* is a statement or expression that evaluates to either TRUE or FALSE. These statements are generally used to help a program determine whether it should (or shouldn't) do (or continue doing) some particular set of instructions."
  ))
})
```

The form of an `if` statement in R is:

```{r, eval=FALSE, echo=TRUE}
if (conditional) {
  # code to run if conditional evaluates to TRUE
} else if (second conditional) {
  # code to run if second conditional evaluates to TRUE
} else {
  # code to run if all conditionals evaluate to FALSE
}
```

The only part actually *required* is the `if (conditional) {}`, all the 
`else if (conditional) {}` and `else {}` blocks are optional. Unlike some other 
programming languages you may come across, R allows you to assign the result of
an `if` statement to a variable directly, like so:

```{r, echo=TRUE, eval=FALSE}
favorite_pet <- if (allergic_to_cats) {
  "Dog"
} else {
  "Cat"
}
```

```{r, echo=TRUE, eval=FALSE}
favorite_pet <- if (allergic_to_cats) "Dog" else "Cat"
```

If you put the entire `if/else` expression on one line, you can omit the `{}`,
otherwise you need them to define a **code block**, or, a set of statements that
are grouped together. If you try to assign the value of an `if` statement without
an `else` clause and the conditional evaluates to `FALSE`, then the assigned 
value will be `NULL`. Also, be aware, if your *conditional* expression does not
evaluate to a logical value (say `1 + 1`), you will get an error message. It's 
also important that *conditional* evaluates to a *single* logical value. A 
*conditional* that evaluates to `[TRUE, FALSE, FALSE]` will act as if it evaluated
to just `TRUE` in an `if` statement, and the `FALSE` values will just be ignored.
R will issue a warning message in these cases, but it will not stop program 
execution.

#### `if` Statement Practice

Practice with the following `if/else` statements. Can you get it to produce the 
"Something strange is happening..." result?

```{r exercise_four, exercise=TRUE, exercise.lines = 12, exercise.eval = T}
x <- 1
y <- 2

if (x == y) {
  "'x' and 'y' are equal"
} else if (x > y) {
  "'x' is greater than 'y'"
} else if (x < y) {
  "'x' is less than 'y'"
} else {
  "Something strange is happening..."
}
```

### Choices - `ifelse()`

The second option, the `ifelse()` function, operates similarly to an `if` statement.
The full method signature is `ifelse(test, yes, no)` where `test` should be a
value or values that can be converted to logical (via `as.logical()`), `yes` 
contains the value or values to return if `test` is `TRUE`, and `no` contains the
value or values to return if `test` is `FALSE`. The biggest difference between
`ifelse()` and an `if` statement is that `ifelse()` can handle multiple 
*conditionals* at the same time. For example:

```{r echo=TRUE}
test <-  c(TRUE, FALSE, TRUE, NA)
yes <-  c(2, 4, 6, 8)
no <-  c(1, 3, 5, 7)
ifelse(test, yes, no)
```

<small>Don't worry too much about that `c()` function, for now just consider 
`c(TRUE, FALSE, TRUE)` to be equivalent to a list of three values: `TRUE`, `FALSE`,
and `TRUE`.</small>

We see that `ifelse()` here returns the first value of `yes`, the second value
of `no`, and the third value of `yes`. The fourth value returned was `NA`, because
the fourth value of `test` was `NA`. In general, it is best if `yes` and `no`
are of the same type, but this isn't strictly required. If `yes` or `no` has
fewer values than `test`, `ifelse()` will 'recycle' values (start over from the
beginning) from `yes` or `no` to return as many values as are in `test`. This
can be useful if you want to have a single value for instances when `test` is
`FALSE`, for example:

```{r echo=TRUE}
test <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
yes <- c(1, 2, 3, 4, 5)
ifelse(test, yes, no = NA)
```

#### `ifelse()` Practice

Practice with the following `ifelse()` function call. Can you adjust the `test`
values to yield a sum of `42`?

```{r exercise_five, exercise=TRUE, exercise.lines = 5, exercise.eval = T}
test <- c(TRUE, FALSE, FALSE, TRUE, FALSE, TRUE, NA, FALSE)
yes <- c(1, 2, 3, 4, 5, 6, 7, 8)
no <- c(0, 2, 4, 6, 8, 10, 12, 14)
result <- ifelse(test, yes, no)
sum(result)
```

### Choices - `switch()`

The final way to make a choice about program flow is the `switch()` function.
`switch()` essentially serves as a more compact replacement of an `if` statement
with multiple `else if` blocks, like this:

```{r echo=TRUE}
animal <- 'fox'
animal_sound <- switch(
  animal,
  dog = 'woof', 
  cat = 'meow',
  bird = 'tweet',
  mouse = 'squeek',
  warning(paste('What does the', animal, 'say?'))
)
```

The first argument passed to `switch()` should either evaluate to a single number 
or a single character string, and the remaining arguments (except for the last) 
should be named arguments that represent `input = output` pairs, where `input` is 
a potential match for the first argument to `switch()` and `output` is the return 
value if `input` matches. The final argument represents a default value. Often, 
this last argument will throw an error, stopping execution if none of the options
match. Regardless, you should include a default argument, otherwise `switch()` 
will return `NULL` as the default value in case no matches are found.

#### `switch()` Practice

Practice with the following `switch()` function call. What animal says 'honk'?

```{r exercise_six, exercise=TRUE, exercise.lines = 10, exercise.eval = T}
animal <- 'fox'
animal_sound <- switch(
  animal,
  dog = 'woof', 
  cat = 'meow',
  bird = 'tweet',
  mouse = 'squeek',
  warning(paste('What does the', animal, 'say?'))
)
```

### Loops - `for`

In general, a **loop** is a programming structure that repeats a sequence of 
statements or expressions either a certain number of times or until a certain
condition is met. `for` loops are used to `r def_link("iterate")` over a list 
of values. The syntax for a for loop is:

```{r def_iterate, context="server"}
observeEvent(input$`iterate`, {
  showModal(modalDialog(
    title = "Iteration",
    "In programming, we use the term *iteration* to refer to both the act of repeating a set of instructions multiple times and a single execution of that set of instructions. You may 'iterate' over a set of numbers, for example, completing the same set of operations while only changing the value of the number operated on. You can think of an iteration as a loop. Wash, rinse, repeat."
  ))
})
```

```{r, eval=FALSE, echo=TRUE}
for (item in list) do_something(item)
```

```{r, eval=FALSE, echo=TRUE}
for (item in list) {
  do_something()
  do_something_else()
  finish(item)
}
```

Just as with `if` statements, the `{}` are optional if executing only a single
statement on one line. Unlike an `if` statement, the `for` loop cannot be assigned
to a value and does not return anything. In a loop (including `for` loops), there
are two special keywords in R that can modify the loop: `next` and `break`.
`next` will cause the loop to skip the rest of the current iteration and move on
to the next iteration. `break` will cause the loop to end prematurely, completing
**no** further iterations.

```{r echo=TRUE}
numbers <- 1:100  # Range from 1 - 100
for (number in numbers) {
  # If the number is 55
  if (number == 55) break
  
  # If the number is not divisible by 10
  if (number %% 10 > 0) next
  
  print(number)
}
```

#### `for` Loop Practice

Practice with the following `for` loop. Can you use `next` and `break` to print
out all the prime numbers less than 100?

```{r exercise_seven, exercise=TRUE, exercise.lines = 20, exercise.eval = F}
numbers <- 1:1000
already_reviewed <- vector("logical", 1000)
for (number in numbers) {
  # These two lines check to see if the number is prime, and mark that
  # number as already reviewed. `1` is ignored, since every number is 
  # evenly divisible by `1`.
  is_prime <- number == 1 | all(number %% numbers[already_reviewed] > 0)
  already_reviewed[number] <- number > 1  # set to TRUE if number > 1
  
  print(number)
}
```

### Loops - `while`

While a `for` loop iterates a number of times that may be known ahead of time, 
a `while` loop instead iterates until *some condition is fulfilled*. The syntax
for a `while` loop is:

```{r, echo=TRUE, eval=FALSE}
while (conditional) {
  do_something()
  conditional <- update_condition()
}
```

A `while` loop will iterate forever (creating an *infinite loop*) unless and until
*conditional* evaluates to `FALSE`. You can still use `next` and `break` within
a `while` loop to skip an iteration or break out of the loop early. In practice,
unless you're absolutely certain your *conditional* will definitely evaluate to
`FALSE` in a reasonable amount of time (you get to decide what is reasonable), you
can always include a built-in 'cut-off', like so:

```{r, echo=TRUE, eval=FALSE}
remaining_attempts <- 100
item <- NULL

# This loop will run until either `item` is something other than NULL, OR
# the number of remaining attempts is zero
while (is.null(item) & remaining_attempts > 0) {
  item <- search_for_item()
  remaining_attempts <- remaining_attempts - 1
}
```

Anything you can do with a `for` loop, you can do with a `while` loop, but many
times it is more convenient to write the `for` loop and fall back to a `while` 
loop if you need the additional flexibility of not deciding ahead of time how
many iterations to expect.

#### `while` Loop Practice

Practice with the following `while` loop. Try to get the biggest number you can
in a 'reasonable' amount of time?

```{r exercise_eight, exercise=TRUE, exercise.lines = 10, exercise.eval = F}
random_number <- 0
biggest_number <- 0
while (random_number < 50) {
  random_number <- runif(1, 0, 100)  # One random number from 0 - 100
  if (random_number > biggest_number) biggest_number <- random_number
  cat('.')  # Prints without a new line between each print
}
print(biggest_number)
```

### Loops - `repeat`

The final type of loop, `repeat`, is an infinite loop by default that you can
only exit with a `break` command. As such, the syntax is even simpler:

```{r, echo=TRUE, eval=FALSE}
repeat {
  do_something()
}
```

This loosely translates to "do a thing over and over, forever". Here's a `repeat`
example that prints out the first 5 Fibonacci numbers (in a bit of a convoluted
way).

```{r echo=TRUE}
fibonacci_numbers <- vector("numeric", 5)
nth_fib <- 0
number_to_try <- 1

repeat {
  if (nth_fib >= 5) break  # Break if we've already found five numbers
  
  # The first two Fibonacci numbers (other than 0) are 1 and 2. Plus, our 
  # `fibonacci_numbers` won't have anything in it yet until we've added 1 and 2
  next_fib <- if (number_to_try %in% c(1, 2)) {
    number_to_try
  } else {
    # The next Fibonacci number is the sum of the last two
    fibonacci_numbers[nth_fib] + fibonacci_numbers[nth_fib - 1]
  }
  
  # If the number we're trying is equal to the next expected Fibonacci number
  # put that number in the `fibonacci_number`s
  if (number_to_try == next_fib) {
    fibonacci_numbers[nth_fib + 1] <- number_to_try
    nth_fib <- nth_fib + 1
  }
  
  # Increase the `number_to_try` by one to try the next number
  number_to_try <- number_to_try + 1
}

print(fibonacci_numbers)
```

So, you can see, while `repeat` is simple on its own, you need to manually write
more code to check a condition and decide whether or not to break from the loop.
You can use a `repeat` to do anything that a `while` or `for` loop can do, but again,
it's often more convenient to use a `for` or `while` when you can, and only drop
back to `repeat` when you really need it. Mostly, you won't **need** it.

#### `repeat`` Loop Practice

Practice with the following `repeat` loop. See if you can get it to print an actual
word (and then stop).

```{r exercise_nine, exercise=TRUE, exercise.lines = 15, exercise.eval = F}
random_string <- ""
current_letter <- 1
repeat {
  random_letter <- sample(letters, 1)
  random_string[current_letter] <- random_letter
  current_letter <- current_letter + 1
}
```










