---
title: "R Data Structures"
output: 
  learnr::tutorial:
    theme:
      version: 3
      bootswatch: sandstone
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(bslib)

if (Sys.getenv("CONTAINERIZED") == "TRUE") {
  source("/usr/bin/learnr/util.R")
} else {
  source(rprojroot::is_rstudio_project$find_file("src/learnr/R/util.R"))
}
knitr::opts_chunk$set(echo = FALSE)
```


## Data Structures {#data-structures}

Let's be honest, if you needed to process every piece of information one at a
time in R, you'd be better off using a spreadsheet. While `x <- 5` is
...exciting..., working with single values at a time is a bit limiting. That's
where `r def_link('data structures')` come in handy. In R, there are a
number of basic data structures that you'll use more or less often, including
**vectors** (often), **matrices** (less often), **lists** (often), **factors**
(less often), and **dataframes** (very often). There are also **arrays**, but we
won't talk about them, because, by the time you actually need to use one, you'll
be far beyond this introductory tutorial. Don't worry, though, we'll walk
through each of the other data structures. In fact, we've actually been using a
data structure this whole time without telling you, the **vector**.

```{r defs1, context="server"}
observeEvent(input$`data-structures`, {
  showModal(modalDialog(
    title = "Data Structures",
    "A data organization, management, and storage format that enables efficient access and modification. More precisely, a data structure is a collection of data values, the relationships among them, and the functions or operations that can be applied to the data."
  ))
})
```

## Vectors

All of our variables have been vectors up to now. In R, a **vector** is a data
structure that consists of a sequence of values that are all of the same type. A
'single' value, like `1`, is just a vector that is one item long. A vector can
contain any number of values, so long as they are all the same type. These are
also referred to as *atomic* vectors in a number of places (including
documentation and error messages). This is to distinguish vectors from another
data structure called a **list**, which can also be called a *generic* vector.
For this tutorial, when I mean *atomic vector*, I'll use the term *vector*, and
when I mean *generic vector*, I'll use the term *list*.

Vectors have three properties accessed through three different functions, one of
which you already know:

-   A type, accessed through `typeof()`
-   A length, accessed through `length()`
-   Additional attributes, accessed through `attributes()`

Vectors can be constructed using the concatenate, `c()`, function. Vectors are
super powerful because you can perform operations on *every* value in a vector
at the same time, as shown in the example below:

```{r echo=TRUE}
example_vector <- c(1, 2, 3, 4, 5, 6)
example_vector^2  # Raise all values to the power of 2
```

<br/>If you attempt to assign values of different types to a single vector, all
values will be automatically converted to the most flexible type, like so:

```{r echo=TRUE}
c(1, FALSE)
```

```{r echo=TRUE}
c('a', 1, FALSE)
```

<br/>Vectors can be unnamed (like the ones we've created so far) or named.

```{r echo=TRUE}
c(first_one = 1, second_one = 2, third_one = 3)
```

<br/>Individual values from vectors can be accessed using either their
[**index**](#index) or their name, if the vector is named.

```{r echo=TRUE}
x <- c(first_one = 1, second_one = 2, third_one = 3)
print(x[2])
print(x['second_one'])
```

<br/>Finally, vector math allows you to use vectors as arguments to any
operation, on a value-by-value basis. So, if you add `c(1, 2)` and `c(3, 4)`,
for example, you get `c(4, 6)`, which is `1 + 3` and `2 + 4`. Some more examples
include:

```{r echo=TRUE}
first_one <- c(10, 10, 10)
second_one <- c(5, 10, 15)
first_one - second_one
```

```{r echo=TRUE}
first_one * second_one
```

```{r echo=TRUE}
first_one / second_one
```

<small><i>Notice how the output from that last example includes decimals?
Remember, a vector contains elements that are all the same type, so when an
operation on one set of elements, `10/15`, returns a decimal, the entire vector
is represented as a `double`.</i></small>

Many (most) functions included in the box with R operate on multiple values in
vectors by default, as well, sometimes returning a single value (or a vector
containing a single item) and sometimes returning a vector with multiple values,
depending on the function. Here are some examples:

```{r echo=TRUE}
x <- c(2, -4, 6, -8, 10, NA)
max(x, na.rm = TRUE)  # Returns the maximum value in the vector
```

```{r echo=TRUE}
min(x, na.rm = TRUE)  # Returns the minimum value in the vector
```

```{r echo=TRUE}
mean(x, na.rm = TRUE)  # Returns the average of the values
```

```{r echo=TRUE}
abs(x)  # Converts all values to absolute value
```

```{r echo=TRUE}
as.character(x)  # Converts all values to 'character' types, if possible
```

```{r echo=TRUE}
is.na(x)  # Determines which values are missing, if any
```

<small><i>Note the additional argument to the `max`, `min`, and `mean`
functions. By default, passing a vector containing an `NA`, or missing, value to
these functions will yield an `NA` as the result. The `na.rm = TRUE` argument
instructs the function to ignore the missing values and return an answer as if
the explicit missing values weren't there.</i></small>

### Quiz

*Try to answer these questions about vectors. Feel free to look back if you need
to.*

```{r vector_questions}
quiz(
  question(
    'Which of the following statments is true about vectors?',
    answer(
      'Vectors always contain multiple values.',
      message = paste(
        'Vectors can also contain just a single value. In fact, every',
        'single value is really a vector in disguise.\n'
      )
    ),
    answer('Values in a vector must be of the same data type.', correct = T),
    answer('All vectors have a `type` and a `length`.', correct = T),
    answer(
      'Every element in a vector must have a name.',
      message = paste(
        'Vector elements may be nammed or unnamed. In fact, you can have a',
        'vector where only **some** elements are named.\n'
      )
    ),
    answer('You can access individual values in a vector by index or by name', correct = T),
    answer(
      'You can perform the same operations on vectors as you can on single values.', 
      correct = T
    ),
    answer(
      'When we say "vector", we are referencing "generic vectors" usually.',
      message = 'The term "vector" generally refers to "atomic vectors".\n'
    ),
    allow_retry = T,
    random_answer_order = T
  ),
  question(
    paste(
      'What is the **output** of this **operation**?',
      '```',
      '  x <- c(1, 2, 3, 4)',
      '  y <- c(4, 3, 2, 1)',
      '  x + y',
      '```',
      sep = '\n'
    ),
    answer('c(5, 5, 5, 5)', correct = T),
    answer('c(1, 2, 3, 4, 4, 3, 2, 1)'),
    answer('20'),
    incorrect = 'Remember, operations on vectors work on an element-by-element basis.',
    allow_retry = T,
    random_answer_order = T
  )
)
```

## Matrices

In R, a **matrix** has absolutely nothing to do with spoons. Instead, matrices build
on the concepts introduced by vectors by representing *2-dimensional* vectors.
You can picture a matrix as a 2D, rectangular grid containing items all of the
same type. A matrix can be constructed using the `matrix()` function, which has
the following [**method signature**](method-signature):

    matrix(data = NA, nrow = 1, ncol = 1, byrow = FALSE, dimnames = NULL)

The arguments accepted by the `matrix()` function are as follows:

-   **data**: an optional data vector (or list). Lists are cast to vectors via
    `as.vector()`
-   **nrow**: the desired number of rows.
-   **ncol**: the desired number of columns.
-   **byrow**: logical. If FALSE (the default) the matrix is filled
    column-by-column, otherwise the matrix is filled by row-by-row.
-   **dimnames**: A "dimnames" attribute for the matrix: NULL or a list of
    length 2 giving the row and column names respectively. An empty list is
    treated as NULL, and a list of length one as row names. The list can be
    named, and the list names will be used as names for the dimensions.

<small><i>If you're using RStudio, this information can be accessed directly in
the console by typing "?matrix" and hitting return. In fact, the documentation
for most packages and functions can be accessed this way. This is EXTREMELY
helpful, even for really experienced R developers.</i></small>

Note that the `data` argument is optional, meaning you can create an empty
matrix. By default, this function will turn a one-dimensional vector into a
two-dimensional matrix, filling in values from `data` column-by-column to yield
the desired number of rows/columns. `dimnames` can be used to name the rows and
columns, much like naming the elements in a vector. Also, you can get away with
specifying only `nrow` or `ncol`, given one, R will calculate the other for you
based on the length of the `data` argument. Here's a concrete example:

```{r echo=TRUE}
y_names <- paste("y", c(1:5), sep = "")
x_names <- paste("x", c(1:20), sep = "")
matrix(c(1:100), nrow = 5, ncol = 20, dimnames = list(y = y_names, x = x_names))
```

</br>You can see that the numbers filled up the first column completely before
beginning to fill the second column.

You can access matrix elements using indices or names, much like with vectors.
For the example above, if we assigned the matrix to the variable `m`, we could
access the value on the third row, fifth column as either `m[3,5]` or
`m["y3","x5"]`. Of interest, you can also access a row or column from the matrix
and treat it as a vector. To get the second row, you would use `m[2, ]` or
`m["y2", ]`. To access the tenth column, you would use `m[ ,10]` or
`m[ ,"x10"]`. The spaces there aren't actually necessary, but they make it
easier to read.

Usefully, you can perform vector math operations on these parts of matrices just
like you can on regular vectors, such as `m[1, ] + m[3, ]` to add the first and
third rows. You can also perform vector math on matrices directly, just like
with vectors:

```{r echo=TRUE}
matrix1 <- matrix(c(1:4), nrow = 2, ncol = 2)
matrix2 <- matrix(c(5:8), nrow = 2, ncol = 2)
matrix1 + matrix2
```

```{r echo=TRUE}
matrix2 / matrix1
```

```{r echo=TRUE}
max(matrix1)
```

You can select a 'sub-matrix', i.e. multiple rows and columns, by using
ranges in your index operation instead of just plain numbers. For example, 
`m[1:3, 1:3]` would return a 3x3 sub-matrix of `m`, starting at the first row,
first column and selecting to the third row, third column.


Matrices also support the `typeof()`, `length()`, and `attributes()`
functions. If you pass a matrix to `attributes()`, one of the attributes
returned will always be `dim`, or dimensions, giving you the number of rows and
columns in the matrix. You can also get the dimensions of a matrix with the
`dim()` function, and the names of the dimensions (if you named them) with the
`dimnames()` function.

```{r echo=TRUE}
matrix3 <- matrix(
  c(1:9), 
  nrow = 3, 
  dimnames = list(foo = c('a', 'b', 'c'), bar = c('x', 'y', 'z'))
)
matrix3
```

```{r echo=TRUE}
dim(matrix3)
```

```{r echo=TRUE}
dimnames(matrix3)
```

### Quiz

*Try to answer these questions about matrices. Feel free to look back if you
need to.*

```{r matrix_questions}
quiz(
  question(
    'Which of the following statments is true about matrices?',
    answer(
      'A matrix must have multiple rows and columns.',
      message = paste(
        'Not necessarily. Your matrix can be 1x1, just line a vector can',
        'have a length of one, or it can be "one-dimensional" with multiple',
        'rows and a single column, or vice versa.\n'
      )
    ),
    answer('Values in a matrix must be of the same data type.', correct = T),
    answer(
      'All matrices have a `dimnames` attribute.', 
      message = paste(
        'Named rows and columns are optional for matrices, just like named',
        'elements are optional for vectors. Try `attributes()` on a matrix to',
        'see.\n'
      )
    ),
    answer('All matrices have a `dim` attribute.', correct = T),
    answer('You can access individual values, rows, or columns in a matrix by index or by name', correct = T),
    answer(
      'You can perform the same operations on matrices as you can on single values or vectors.', 
      correct = T
    ),
    answer(
      'You can index a matrix like so: `m[column_number, row_number]`',
      message = paste(
        'One of the tricky things to get used to is the fact that, in most',
        'programming languages, 2D data is referenced by',
        '[row_number, column_number], when you may be used to thinking of',
        'these locations as (x, y) coordinates.\n'
      )
    ),
    allow_retry = T,
    random_answer_order = T
  ),
  question(
    paste(
      'What is the **output** of this **operation**?',
      '```',
      'matrix(c(1:9), nrow = 3, byrow = T)',
      '```',
      sep = '\n'
    ),
    answer(
      paste(
        '```',
        '     [,1] [,2] [,3]',
        '[1,]    1    2    3',
        '[2,]    4    5    6',
        '[3,]    7    8    9',
        '```',
        sep = '\n'
      ), 
      correct = T
    ),
    answer(
      paste(
        '```',
        '     [,1] [,2] [,3]',
        '[1,]    1    4    7',
        '[2,]    2    5    8',
        '[3,]    3    6    9',
        '```',
        sep = '\n'
      )
    ),
    answer(
      paste(
        '```',
        '     [,1] [,2] [,3]',
        '[1,]    7    8    9',
        '[2,]    4    5    6',
        '[3,]    1    2    3',
        '```',
        sep = '\n'
      )
    ),
    incorrect = paste(
      'The `byrow = FALSE` argument means the matrix will fill up the first row',
      'before adding numbers to the second row.'
    ),
    allow_retry = T,
    random_answer_order = T
  )
)
```

## Lists

What if you could put *different* types of things in a vector? Well, then you'd
have that *generic vector* we mentioned in the section on vectors, better 
known as a **list**. Not only can lists contain multiple values of different
types, they can contain other data structures, too. So, you can put a vector
in a list, a matrix in a list, a list in a list in a list in a list (you get
the idea). You create a list with the `list()` function, passing named or unnamed
arguments as the new list elements.

```{r echo=TRUE}
a_vector <- c(1:10)
a_matrix <- c(1:9) * 11
a_word <- 'word'
a_list <- list(1, 'two', c(1, 1, 1))
list(a_vector = a_vector, a_matrix = a_matrix, a_word = a_word, a_list = a_list)
```

</br>Now, there are a few things to note about accessing elements in a list. To begin,
`a_list[1]` and `a_list[[1]]` are different things. The first version, with only
one set of brackets, returns a *list* containing the first item from `a_list`, 
while the second version with double brackets returns the *value* of the first
item from `a_list`. You can confirm this by:

```{r echo=TRUE}
typeof(a_list[1])
```

```{r echo=TRUE}
typeof(a_list[[1]])
```

</br>You can still access list items by name, using the `a_list["name"]` and
`a_list[["name"]]` syntax, similar to vectors. but now you have another option:
`a_list$name`. This explains the `$a_vector`, `$a_matrix`, `$a_word`, and 
`$a_list` you saw in the output for our example list, these were the named
items in the list. Using `a_list$name` is equivalent to `a_list[["name"]]`, just
with fewer characters.

While lists give you a lot of flexibility, you also lose some functionality going
from a vector to a list, namely the vector math. You can no longer add, subtract,
get the maximum value, or get the average of a list, because it might not
contain items that make sense for those functions. After all, what is the average
value of [1, "tacos", [5, 6, 7]]? If you know your list contains only items
of the same type (or items that can be converted to the type you want), you can
use the `unlist()` function to `r def_link("flatten")` your list into a vector.

```{r echo=TRUE}
another_list <- list(1, "tacos", c(5, 6, 7))
unlist(another_list)
```

</br>You can see that we got a character vector back, since that was the most
flexible data type in our list (refer back to our section on vectors if you
find this behavior to be strange).

You can get more information about a list using our `typeof()`, `length()`, and
`attributes()` functions. You can `r def_link("append")` items to a list using
`a_list[[length(a_list) + 1]] <- new_value` (admittedly a bit cumbersome) or 
insert named elements with `a_list$new_element <- new_value` (better). You can
replace a list items value with `a_list$element_name <- new_value`. You can
also remove an item from a list using `a_list$element_to_remove <- NULL`, i.e.
by replacing the list item's value with `NULL`.

```{r defs2, context="server"}
observeEvent(input$`flatten`, {
  showModal(modalDialog(
    title = "Flatten",
    "A verb that refers to the act of taking a nested data structure, like a list containing lists where each list contains more lists, and turning it into a list that contains only \"single\" elements (like numbers, strings, objects, etc.)."
  ))
})

observeEvent(input$`append`, {
  showModal(modalDialog(
    title = "Append",
    "A verb that refers to the act of adding one more item to a data structure, typically to the \"end\" of that data structure."
  ))
})
```

### Quiz

*Try to answer these questions about lists. Feel free to look back if you
need to.*

```{r list_questions}
quiz(
  question(
    'Which of the following statments is true about lists?',
    answer(
      'You can access the first value in a list with `a_list[1]`.',
      message = paste(
        'This is one of the more potentially confusing things about lists,',
        'accessing items using single brackets "[]" yields a *list* containing',
        'that item, using double brackets "[[]]" yields the item itself.\n'
      )
    ),
    answer(
      'Values in a list must be of the same data type.', 
      message = paste(
        'Unlike a vector, items in a list do not need to be the same type, and',
        'they can be other data structures in addition to primitive types.\n'
      )
    ),
    answer('You can access a list item by name using `a_list$item_name`', correct = T),
    answer('You can convert a list to a vector with the `unlist()` function.', correct = T),
    answer('You can remove an item from a list by replacing it with `NULL`', correct = T),
    answer(
      'You can perform the same operations on lists as you can on single values or vectors.', 
      message = paste(
        'Lists do not support vectorized math operations the way single data type',
        'data structures do.\n'
      )
    ),
    answer(
      'To create a new list, pass list items as arguments to the `list()` function',
      correct = T
    ),
    allow_retry = T,
    random_answer_order = T
  ),
  question(
    paste(
      'What is the **output** of this **operation**?',
      '```',
      'a_list <- list(TRUE, FALSE, matrix(c(1:9), nrow = 3))',
      'unlist(a_list)',
      '```',
      sep = '\n'
    ),
    answer('`1 0 1 2 3 4 5 6 7 8 9`', correct = T),
    answer('`TRUE FALSE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE`'),
    answer('`"TRUE" "FALSE" "1" "2" "3" "4" "5" "6" "7" "8" "9"`'),
    incorrect = paste(
      'The resulting vector will default to the most flexible data type',
      '*in the list* when converting a list to a vector.'
    ),
    allow_retry = T,
    random_answer_order = T
  )
)
```


## Factors

Factors are a bit of a funny data structure. They don't quite follow the logical
progression from vectors to matrices to lists to dataframes (we'll get to those
next). Factors are very useful, however, for representing *categories*. Let's
say, for example, that you have a character vector containing the colors of the
rainbow as strings. Now, this vector has millions of these color names, and 
you'd like to:

1. Reduce the size of that vector in memory.
2. Speed up operations on that vector.
3. Sort that vector some way other than alphabetical.
4. Restrict strings in that vector to the list of colors you already have.

You can do all of that with a **factor**.

```{r echo=TRUE}
de_colores <- rep(
  c('orange', 'blue', 'indigo', 'red', 'green', 'violet', 'yellow', 'brown'),
  1000000
)
de_colores_factor <- factor(
  de_colores,
  levels = c('red', 'orange', 'yellow', 'green', 'blue', 'indigo', 'violet'),
  labels = c('RED', 'ORANGE', 'YELLOW', 'GREEN', 'BLUE', 'INDIGO', 'VIOLET'),
  ordered = TRUE,
  nmax = 7
)
str(de_colores_factor)
```

```{r echo=TRUE}
head(de_colores_factor, 10)
```

```{r echo=TRUE}
levels(de_colores_factor)
```

```{r echo=TRUE}
attributes(de_colores_factor)
```


In testing, `de_colores` above was 64MB, while `de_colores_factor` was 32MB, 
cutting the original vector's size in half. How is this done? Using the `str()`
(short for **str**ucture) function on our factor gives us a hint. An R factor is
really, deep down, a numeric vector where each number is mapped to a label which
is one of the original strings in our vector. Since a number is 'smaller' in 
memory than a string (even a small one), the factor is smaller. (This isn't the
whole story, R does some things to make strings in memory more efficient than
they otherwise would be. Let's save that discussion for later.) Not only that, 
but many operations will be faster because the vector is now really numeric.

Let's take a look at the full set of arguments to `factor()` to learn more about 
what we can do with them:

```
factor(
  x = character(), 
  levels, 
  labels = levels, 
  exclude = NA, 
  ordered = is.ordered(x), 
  nmax = NA
)
```

- **x**: A character vector, or something that can be converted to a character 
vector.
- **levels**: an optional vector of the unique values (as character strings) 
that x might have. The default is the unique set of strings in x, sorted into 
increasing order of x. The resulting factor can be sorted in order of the
levels. Any value in x that isn't represented in **levels** will be replaced
with `NA` in the resulting factor (that's where 'brown' went).
- **labels**: either an optional character vector of labels for the levels 
(in the same order as levels after removing those in exclude), or a character 
string of length 1. Duplicated values in labels can be used to map different 
values of x to the same factor level. Labels can be applied to change how the
factor values are displayed.
- **exclude**: a vector of values to be excluded when forming the set of levels. 
This may be a factor with the same level set as x or it could be a character
vector. This is most useful when you're not specifying **levels** explicitly, to 
remove values from x.
- **ordered**: logical flag to determine if the levels should be regarded as 
ordered (in the order given). Mostly this just modifies the way a factor is 
represented when you print it, but some functions will treat ordered and
unordered factors differently, so get familiar with function documentation. 
Sorting will still be based on the order of the factor levels.
- **nmax**: an upper bound on the number of levels. The actual usage is a bit
arcane, but if you set this to be the number of factor levels you expect, the
conversion into a factor will be more efficient. Not a huge deal to leave it out
in most cases.

As shown above, you can check what levels are in your factor using the `levels()`
function. Our old standbys of `length()`, `type()`, and `attributes()` still 
work, too. In most other ways, you can treat a factor as a normal vector. For
now, the last thing we'll cover is changing the levels.

```{r echo=TRUE}
# To change factor levels, assign directly to levels(f)
levels(de_colores_factor) <- c(
  'rojo', 'naranja', 'amarillo', 'verde', 
  'azul', 'indigo',  'violeta'
)
str(de_colores_factor)
```

```{r echo=TRUE}
head(de_colores_factor, 10)
```

```{r echo=TRUE}
attributes(de_colores_factor)
```

We've 're-leveled' this factor, using different names for the values. Assigning 
to `levels(de_colores_factor)` may look a bit strange, but it works! There's
lots more we can do with factors and some R libraries to make it easier. We'll
cover discuss that later, in the relevant sections.

### Quiz

*Try to answer these questions about factors. Feel free to look back if you
need to.*

```{r factor_questions}
quiz(
  question(
    'Which of the following statments is true about factors?',
    answer(
      paste('When creating a factor, you need to pass the argument',
            '`ordered = TRUE` in order for it to sort properly.'),
      message = paste(
        'A factor is always sorted by the order of its levels. Passing',
        '`ordered = TRUE` just marks the factor as "ordered", which will change',
        'how some functions treat the factor.\n'
      )
    ),
    answer(
      'You can only create a factor from a character vector.', 
      message = paste(
        'A factor can be created from a character vector, or anything that can',
        'be converted to a character vector using `as.character()`\n'
      )
    ),
    answer(
      paste('When creating a factor, you **must** include every string in the',
            'original character vector as one of the levels.'), 
      message = paste(
        'If you do not specify levels, then every string will be considered.',
        'If you do specify levels, then any string you do not explicitly include',
        'will be converted to an `NA` in the resulting factor.'
      )
    ),
    answer('Factors are most useful for representing categorical values.', correct = T),
    answer('You can rename factor levels with `levels(a_factor) <- new_levels`.', correct = T),
    answer(
      'A typical factor will take up about half as much memory as a character vector', 
      correct = T
    ),
    answer(
      paste('You can view factor levels using the `print()`, `attributes()`,',
            'and `levels()` functions'),
      correct = T
    ),
    allow_retry = T,
    random_answer_order = T
  ),
  question(
    paste(
      'In what order will the levels of `a_factor` be?',
      '```',
      'weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"',
      'a_factor <- factor(weekdays)',
      '```',
      sep = '\n'
    ),
    answer('Monday, Tuesday, Wednesday, Thursday, Friday'),
    answer('Friday, Monday, Thursday, Tuesday, Wednesday', correct = T),
    answer('Friday, Thursday, Wednesday, Tuesday, Monday'),
    incorrect = paste(
      'Since the levels are not specified, how do factors sort themselves by',
      'default?'
    ),
    allow_retry = T,
    random_answer_order = T
  )
)
```

## Dataframes

Oh boy! The moment you've been waiting for, even if you didn't know it yet. 
Dataframes are the biggest reason you should choose to learn R for your data
cleaning, wrangling, analyzing, extracting, etc. tasks. In a very real sense,
a dataframe is a list of vectors, where each vector must be the same length
and represents a single column in a data table.

```{r echo=TRUE}
a_dataframe <- data.frame(
  numbers = runif(10, 0, 100),
  dates = seq(Sys.Date(), by = "month", length.out = 10),
  categories = factor(
    sample(rep(c("sm", "md", "lg"), 10), 10),
    levels = c("sm", "md", "lg"),
    labels = c("Small", "Medium", "Large")
  )
)
first_five_rows <- head(a_dataframe, 5)
first_five_rows
```

You can see a summarized set of information about your dataframe using the 
`summary()` function. 

```{r echo=TRUE}
summary(a_dataframe)
```

Or a description of the dataframe's structure with `str()`.

```{r echo=TRUE}
str(a_dataframe)
```

If you are using RStudio (you should be), you can open a viewer for the 
dataframe using the `View(a_dataframe)` function (or clicking on the dataframe 
in the Environment pane). The viewer will allow you to sort by columns, filter
columns, and search the dataframe.

```{r fig1, results="asis"}
insert_figure(
  "view-dataframe-screenshot.png", 
  "RStudio table viewer", 
  caption = "Figure 1: The table viewer built in to RStudio.", 
  max_width = "200px"
)
```

We will spend a *lot* of time with dataframes and the R packages that make
working with dataframes really straightforward, but for now, remember that a 
dataframe is a list with particular constraints. This means that all the 
operations we learned for vectors and lists still apply. You can select a 
particular row from your data frame by index (`df[3, ]`), a range of rows by
index (`df[2:10, ]`), a particular column by index (`df[ ,5]`), multiple
columns by index (`df[ ,6:11]`), a subset of rows and columns by index 
(`df[2:10,6:11]`), and columns by name or heading (`df$age` or `df$start_date`
for example). 

Single columns accessed this way behave just like regular vectors, because they
are regular vectors. They can be summed, added, subtracted, etc.

The column names in a dataframe don't have to comply with typical R variable 
naming conventions, so you can have a column named 'Current Status of Education
Progress' for example. In order to reference columns like this by name, you'll
need backticks (`` `r '\x60'` ``) around the column name, for example 
`` `r 'df$\x60Current Status of Education Progress\x60'` ``. You may find
that to be a pain, but don't worry, you can freely modify column names, but more
on that later.

You can add rows and columns to existing dataframes using the `rbind()` (row bind)
and `cbind()` column bind functions, respectively. When using `rbind()`, the
row or rows you want to add should either be a dataframe with the same columns as
the dataframe you want to add rows to or a list with the same names as your
dataframe columns like so:

```{r echo=TRUE}
a_dataframe <- data.frame(
  numbers = c(1, 2, 3, 4, 5),
  letters = c('a', 'b', 'c', 'd', 'e')
)

b_dataframe <- data.frame(numbers = 6, letters = 'f')

rbind(a_dataframe, b_dataframe)
```

```{r exercise_test_df, exercise=TRUE, exercise.lines = 4, exercise.eval = T}
b_list <- list(numbers = 6, letters = 'f')

rbind(a_dataframe, b_list)
```

```{r exercise_one_df, exercise=TRUE, exercise.lines = 3, exercise.eval = F}
# This one will return an error
c_dataframe <- data.frame(foo = 7, bar = 'g')
rbind(a_dataframe, c_dataframe)
```

`data.frame()` is the base R function for creating a dataframe, but we won't 
spend too much time on it. In fact, we're going to spend most of our time from
now on depending on `r def_link("packages")` designed specifically to make working
with dataframes (and data in general) more powerful and intuitive. So, hang on 
tight, don't touch that dial, all that stuff. Next up, our first foray into **The
Tidyverse**.

```{r defs3, context="server"}
observeEvent(input$`packages`, {
  showModal(modalDialog(
    title = "R Packages",
    "R packages are a collection of R functions, compiled code and sample data. They are stored under a directory called a \"library\" in the R environment. By default, R installs a set of packages during installation. More packages can be installed added later, when they are needed for some specific purpose. The contents of a package can be accessed in your R code by using the `library()` function."))
})
```

